HMMotion: Using tracking data to predict coverage strategies in American Football

### 2024-12-20 submission.Rmd
.

## Introduction

Predicting whether a defense is playing man or zone coverage has long been a focal point in football analytics. A notable example is Amazon's NFL Next Gen Stats model, which delivers predictions during live broadcasts (see a snapshot of the 2024 Week 12 matchup between the Pittsburgh Steelers and Cleveland Browns. These predictions, however, are often limited to analyzing specific plays without incorporating pre-snap motion which is a crucial element of modern offensive strategies.

Our project takes this analysis a step further. While we similarly predict man or zone coverage when the teams are set, we further leverage the additional information of pre-snap player movements. Using hidden Markov models (HMMs), we model defenders' trajectories based on hidden states, representing the offensive players they may be guarding. This probabilistic approach allows us to infer dynamic coverage patterns during plays with pre-snap motion.

To exploit this complex, high-dimensional data into actionable insights, we calculate the entropy of the state probabilities, producing a single measure of uncertainty for each play. Incorporating these as covariates into the existing pre-motion model significantly improves both the AUC and detection accuracy. In doing so, we present a data-driven framework that evaluates the effectiveness of pre-snap motion in uncovering defensive strategies, providing real-time tactical insights for coaches.

>>>>

## Introduction
Deciding between man and zone coverage is one of the most critical strategic choices a defensive coordinator must make before each offensive play in American football. While experienced offensive coordinators and quarterbacks often rely on visual cues to identify these defensive schemes, the increasing availability of player tracking data offers a new avenue to uncover and analyze these tactics. A notable example is Amazon's NFL Next Gen Stats model, which delivers predictions during live broadcasts (see a snapshot of the 2024 Week 12 matchup between the Pittsburgh Steelers and Cleveland Browns. These predictions, however, are often limited to analyzing specific plays without incorporating pre-snap motion which is a crucial element of modern offensive strategies.

Our project takes this model a step further. While we similarly predict man or zone coverage when the teams are set, we further leverage the additional information of pre-snap player movements. Using a hidden Markov model (HMM), we model defenders' trajectories based on hidden states, which represent the offensive players they may be guarding. This probabilistic approach allows us to infer dynamic coverage patterns during plays with pre-snap motion. To exploit this high-dimensional time-series data, we calculate the entropy of the state probabilities, thus, producing measures of uncertainty for each play. Incorporating these as covariates into the existing pre-motion model significantly improves both the AUC and detection accuracy.

In doing so, we present a data-driven framework that evaluates the effectiveness of pre-snap motion in uncovering defensive strategies, providing real-time tactical insights for coaches.
